---
title: "Analysis of Wildlife Strikes to Aircraft"
subtitle: "Practicum I CS5200"
author: "Sai Prasasth Koundinya Gandrakota \\textless{}gandrakota.s@northeastern.edu\\textgreater{}"
email: "gandrakota.s@northeastern.edu"
date: "Spring 2024"
output:
  pdf_document: default

---

## Connecting to Database

```{r Connecting to DB, echo = FALSE, results='hide', warning=FALSE, message=FALSE}
rm(list = ls())

# Required packages
packages <- c("RMySQL", "DBI", "readr", "knitr")

# Install packages if not yet installed
installed_packages <- packages %in% rownames(installed.packages())
if (any(installed_packages == FALSE)) {
  install.packages(packages[!installed_packages])
}

# Load Installed Packages
library(RMySQL)
library(DBI)
library(readr)
library(knitr)

# Connect to database
con <- dbConnect(MySQL(), 
                host = "database-1.czc6oymseewd.us-east-1.rds.amazonaws.com",
                user = "admin",
                password = "8576939257spkg",
                dbname = "WildlifeStrikes",
                port = 3306)

knitr::opts_chunk$set(connection = con) #Setting default connection for all SQL chunks

# Check the connection
dbGetInfo(con)
```

## Defining the DB Schema

```{r DB Schema, echo = FALSE, results='hide', message=FALSE}
# Define the DB Schema

# Drop any existing tables
dbExecute(con, "DROP TABLE IF EXISTS strikes")
dbExecute(con, "DROP TABLE IF EXISTS flights")
dbExecute(con, "DROP TABLE IF EXISTS airports")
dbExecute(con, "DROP TABLE IF EXISTS conditions")


# Create a table that stores airports
# SCHEMA:
#//////////// airports ////////////////
###### aid : INTEGER (PK)
###### airportName : TEXT
###### airportState : TEXT
###### airportCode : VARCHAR(3)
#/////////////////////////////////////
create_airports <- "CREATE TABLE IF NOT EXISTS airports (
                    aid INTEGER AUTO_INCREMENT,
                    airportName TEXT,
                    airportState TEXT,
                    airportCode VARCHAR(3) DEFAULT 'ZZZ',
                    PRIMARY KEY (aid)
                  )"

dbExecute (con, create_airports)


# Create a table flights that stores information about flights that experienced wildlife strikes
# SCHEMA:
#//////////// flights ////////////////
###### fid : INTEGER (PK)
###### date : DATE
###### originAirport : INTEGER (FK)
###### airlineName : TEXT
###### aircraftType : TEXT
###### isHeavy : BOOLEAN
#/////////////////////////////////////
create_flights <- "CREATE TABLE IF NOT EXISTS flights (
                    fid INTEGER AUTO_INCREMENT,
                    date DATE,
                    originAirport INTEGER,
                    airlineName TEXT,
                    aircraftType TEXT,
                    isHeavy BOOLEAN,
                    PRIMARY KEY (fid),
                    FOREIGN KEY (originAirport) REFERENCES airports (aid)
                  )"

dbExecute (con, create_flights)


# Create a lookup table called conditions
# SCHEMA:
#////////// conditions ///////////////
###### cid : INTEGER (PK)
###### sky_condition : TEXT
###### explanation : TEXT
#/////////////////////////////////////
create_conditions <- "CREATE TABLE IF NOT EXISTS conditions (
                    cid INTEGER AUTO_INCREMENT,
                    sky_condition TEXT,
                    explanation TEXT,
                    PRIMARY KEY (cid)
                  )"

dbExecute (con, create_conditions)


# Create a table that stores wildlife strikes called strikes
# SCHEMA:
#////////// strikes //////////////////
###### sid : INTEGER (PK)
###### fid : INTEGER (FK)
###### numBirds : INTEGER
###### impact : TEXT
###### damage : BOOLEAN
###### altitude : INTEGER (>=0)
###### conditions : INTEGER (FK)
#/////////////////////////////////////
create_strikes <- "CREATE TABLE IF NOT EXISTS strikes (
                    sid INTEGER AUTO_INCREMENT,
                    fid INTEGER,
                    numBirds INTEGER DEFAULT 1,
                    impact TEXT,
                    damage BOOLEAN,
                    altitude INTEGER CHECK(altitude >= 0),
                    conditions INTEGER,
                    PRIMARY KEY (sid),
                    FOREIGN KEY (fid) REFERENCES flights (fid),
                    FOREIGN KEY (conditions) REFERENCES conditions(cid)
                  )"

dbExecute (con, create_strikes)
```

## Testing DB Schema

```{r Test DB Schema: 1, echo = FALSE, eval = FALSE}
# TEST CASE: Check data types, primary keys and default values for tables created

# airports table column names, data types, primary keys and default values
dbGetQuery(con, "DESCRIBE airports;")

# flights table column names, data types, primary keys and default values
dbGetQuery(con, "DESCRIBE flights;")

# conditions table column names, data types, primary keys and default values
dbGetQuery(con, "DESCRIBE conditions;")

# strikes table column names, data types, primary keys and default values
dbGetQuery(con, "DESCRIBE strikes;")
```
```{r Test DB Schema: 2, echo = FALSE, eval = FALSE}
# TEST CASE: Check foreign key constraints for flights table

# Insert dummy values into airports table
dbExecute(con, "INSERT INTO airports (airportName, airportState, airportCode)
                VALUES 
                    ('Test Airport 1', 'Test State 1', 'TS1'),
                    ('Test Airport 2', 'Test State 2', 'TS2')
          ")

# Show inserted values in airports table
dbGetQuery(con, "SELECT * FROM airports")

# FAIL CASE (Wrong FK)
# Insert dummy values into flights table 
tryCatch(
  {
    dbExecute(con, "INSERT INTO flights
                    VALUES (1234, '2000-01-01', 4, 'Test Airline', 'Test Aircraft', FALSE)")
  }, 
  error = function(e) {
    print(paste("Error:", e$message))
  }
)

# PASS CASE (Correct FK)
# Insert dummy values into flights table 
dbExecute(con, "INSERT INTO flights
                VALUES 
                    (1234, '2000-01-01', 1, 'Test Airline 1', 'Test Aircraft 1', FALSE),
                    (2131, '2000-12-01', 2, 'Test Airline 2', 'Test Aircraft 2', TRUE)
          ")

# Show inserted values in flights table
dbGetQuery(con, "SELECT * FROM flights")
```
```{r Test DB Schema: 3, echo = FALSE, eval = FALSE}
# TEST CASE: Check foreign key and check constraints for strikes table

# Insert dummy values into conditions table
dbExecute(con, "INSERT INTO conditions (sky_condition, explanation)
                VALUES 
                    ('Test Condition 1', 'Exp 1'),
                    ('Test Condition 2', 'Exp 2')
          ")

# Show inserted values in conditions table
dbGetQuery(con, "SELECT * FROM conditions")

# Show previously inserted values in flights table
dbGetQuery(con, "SELECT * FROM flights")

# FAIL CASE 1 (Wrong FK - fid)
# Insert dummy values into strikes table 
tryCatch(
  {
    dbExecute(con, "INSERT INTO strikes (fid, numBirds, impact, damage, altitude, conditions)
                    VALUES (234, 3, 'Test Impact', FALSE, 1200, 1)")
  }, 
  error = function(e) {
    print(paste("Error:", e$message))
  }
)

# FAIL CASE 2 (Wrong FK - cid)
# Insert dummy values into strikes table 
tryCatch(
  {
    dbExecute(con, "INSERT INTO strikes (fid, numBirds, impact, damage, altitude, conditions)
                    VALUES (1234, 3, 'Test Impact', FALSE, 1200, 5)")
  }, 
  error = function(e) {
    print(paste("Error:", e$message))
  }
)

# FAIL CASE 3 (altitude < 0)
# Insert dummy values into strikes table 
tryCatch(
  {
    dbExecute(con, "INSERT INTO strikes (fid, numBirds, impact, damage, altitude, conditions)
                    VALUES (1234, 3, 'Test Impact', FALSE, -10, 1)")
  }, 
  error = function(e) {
    print(paste("Error:", e$message))
  }
)

# PASS CASE (Correct FK - fid, cid; altitude >= 0)
# Insert dummy values into strikes table 
dbExecute(con, "INSERT INTO strikes (sid, fid, numBirds, impact, damage, altitude, conditions)
                VALUES 
                    (1, 1234, 3, 'Test Impact 1', FALSE, 1000, 1),
                    (2, 2131, 1, 'Test Impact 2', TRUE, 0, 2)
          ")

# Show inserted values in strikes table
dbGetQuery(con, "SELECT * FROM strikes")
```

```{r Deleting test values, echo = FALSE, eval = FALSE}
# Deleting dummy rows from tables
dbExecute(con, "DELETE FROM strikes;")
dbExecute(con, "DELETE FROM flights;")
dbExecute(con, "DELETE FROM conditions;")
dbExecute(con, "DELETE FROM airports;")

# Reset the auto-increment counter for the primary key column in all tables
dbExecute(con, "ALTER TABLE airports AUTO_INCREMENT = 1;")
dbExecute(con, "ALTER TABLE flights AUTO_INCREMENT = 1;")
dbExecute(con, "ALTER TABLE conditions AUTO_INCREMENT = 1;")
dbExecute(con, "ALTER TABLE strikes AUTO_INCREMENT = 1;")
```


## Loading the CSV

```{r Load CSV, echo = FALSE, results='hide', message=FALSE}
# Load the CSV file with data into a data frame
bds.raw <- read_csv("BirdStrikesData-V3.csv")

# Show structure of imported data frame
str(bds.raw)
```

## Preprocessing the data frame

```{r Preprocessing Data, echo = FALSE, message=FALSE}

# Create copy of bds.raw
bds.copy <- bds.raw

# Replace empty strings with NA
bds.copy[bds.copy == ""] <- NA

# Convert all NULL values to default values respective to the columns

# Convert NULL values in airport, origin, sky_conditions, airline, aircraft and impact
# to default value "Unknown"
string_cols <- c("airport", "origin", "sky_conditions", "airline", "aircraft", "impact")
bds.copy[string_cols][is.na(bds.copy[string_cols])] <- "Unknown"

# Convert NULL values in heavy_flag column to "No"
bds.copy[is.na(bds.copy[["heavy_flag"]]), "heavy_flag"] <- "No"

# Convert heavy_flag values to Boolean: Yes = TRUE, No = FALSE
bds.copy$heavy_flag <- ifelse(bds.copy$heavy_flag == "Yes", 1, 0)

# Convert NULL values in damage column to "No damage"
bds.copy[is.na(bds.copy[["damage"]]), "damage"] <- "No damage"

# Convert damaged values to boolean: (Damage, Loss) = TRUE, No Damage = FALSE
bds.copy$damage <- ifelse(bds.copy$damage == "Damage" | bds.copy$damage == "Loss", 1, 0)

# Convert NULL values in altitude column to 0
bds.copy[is.na(bds.copy[["altitude_ft"]]), "altitude_ft"] <- 0

# Remove the separating commas and convert altitude_ft values to numeric values
bds.copy$altitude_ft <- as.numeric(gsub(",", "", bds.copy$altitude_ft))

# Remove rows with negative altitude values
bds.copy <- bds.copy[bds.copy["altitude_ft"] >= 0, ]

# Convert flight_date values to YYYY-MM-DD format
bds.copy$flight_date <- as.Date(bds.copy$flight_date, format = "%m/%d/%Y %H:%M")

# Convert NULL flight_date values to lower bound date - 1900-01-01
bds.copy[is.na(bds.copy[["flight_date"]]), "flight_date"] <- as.Date("1900-01-01")

```

## Importing Data into airports table

```{r Load Airports, results='hide', echo = FALSE, message=FALSE}
# Import data into the airports table

# Select distinct values of required columns from csv file
airports.raw <- unique(bds.copy[, c("airport", "origin")])

# Change column names to match airports table in database
colnames(airports.raw) <- c("airportName", "airportState")

# Write airportName and airportState Columns into the airports table
dbWriteTable(con, name="airports", airports.raw, append=TRUE, row.names = FALSE)

# Store updated airports table in data frame
airports.updated <- dbGetQuery(con, "SELECT * FROM airports")

```

## Importing Data into conditions table

```{r Load conditions, results='hide', echo = FALSE, message=FALSE}
# Import data into the conditions table

# Select distinct values of required columns from csv file
conditions.raw <- unique(bds.copy["sky_conditions"])

# Change column name to match conditions table in database
colnames(conditions.raw) <- "sky_condition"

# Write sky_condition column into the conditions table
dbWriteTable(con, name="conditions", conditions.raw, append=TRUE, row.names = FALSE)

# Store updated conditions table in data frame
conditions.updated <- dbGetQuery(con, "SELECT * FROM conditions")
```

## Importing Data into flights table

```{r Load flights, results='hide', echo = FALSE, message=FALSE}
# Import data into the flights table

# Select values of required columns from csv file
flights.raw <- bds.copy[,c("rid", "flight_date", "aircraft", "airline", "heavy_flag", "airport", "origin")]

# Initialize a vector to store airport ids
airport_ids <- integer(nrow(flights.raw))

#Iterate through each row in flights.raw data frame
for (i in 1:nrow(flights.raw))
{
  # Find index in airport.updated data frame which matches the airport and origin information for current flight row
  airport_match <- which(airports.updated$airportName == flights.raw$airport[i] & airports.updated$airportState == flights.raw$origin[i])
  
  #Check if match exists (If length of match index is 0 then no match is found)
  if (length(airport_match) > 0)
  {
    # If match exists: Store corresponding aid in vector
    airport_ids[i] <- airports.updated$aid[airport_match]
  }
  else
  {
    # If match doesn't exist: Store NA value in vector
    airport_ids[i] <- NA
  }
}

# Drop airport and origin columns from flight.raw data frame
flights.raw = subset(flights.raw, select = -c(airport, origin))

# Add airport_ids column to flights.raw containing the airport_ids vector
flights.raw$airport_ids <- airport_ids

# Rearrange columns in flights.raw to match flights table in database
flights.raw <- flights.raw[,c("rid", "flight_date", "airport_ids", "airline", "aircraft", "heavy_flag")]

# Change column names to match flights table in database
colnames(flights.raw) <- c("fid", "date", "originAirport", "airlineName", "aircraftType", "isHeavy")

# Write fid, date, originAirport, airlineName,aircraftType and isHeavy Columns into the airports table
dbWriteTable(con, name="flights", flights.raw, append=TRUE, row.names = FALSE)

# Store updated table after import
flights.updated <- dbGetQuery(con, "SELECT * FROM flights")
```

## Importing Data into strikes table

```{r Load strikes, results='hide', echo = FALSE, message=FALSE}
# Import data into the strikes table

# Select values of required columns from csv file
strikes.raw <- bds.copy[, c("impact", "damage", "altitude_ft", "rid", "sky_conditions" )]

# Initialize a vector to store flight ids
flight_ids <- integer(nrow(strikes.raw))

# Initialize a vector to store condition ids
condition_ids <- integer(nrow(strikes.raw))

#Iterate through each row in flights.raw data frame
for (i in 1:nrow(strikes.raw))
{
  # Find index in airport.updated data frame which matches the airport and origin information for current strikes row
  flight_match <- which(flights.updated$fid == strikes.raw$rid[i])
  
  # Find index in conditions.updated data frame which matches the sky_conditions information for current strikes row
  conditions_match <- which(conditions.updated$sky_condition == strikes.raw$sky_conditions[i])
  
  # Check if match exists for flights (If length of match index is 0 then no match is found) 
  if (length(flight_match) > 0)
  {
    # If match exists: Store corresponding fid in vector
    flight_ids[i] <- flights.updated$fid[flight_match]
  }
  else
  {
    # If match doesn't exist: Store NA value in vector
    flight_ids[i] <- NA
  }
  
  # Check if match exists for conditions (If length of match index is 0 then no match is found) 
  if (length(conditions_match) > 0)
  {
    # If match exists: Store corresponding cid in vector
    condition_ids[i] <- conditions.updated$cid[conditions_match]
  }
  else
  {
    # If match doesn't exist: Store NA value in vector
    condition_ids[i] <- NA
  }
}

# Drop rid and sky_conditions columns from strikes.raw data frame
strikes.raw = subset(strikes.raw, select = -c(rid, sky_conditions))

# Add flight_ids column to strikes.raw containing the flight_ids vector
strikes.raw$flight_ids <- flight_ids

# Add condition_ids column to strikes.raw containing the condition_ids vector
strikes.raw$condition_ids <- condition_ids

# Rearrange columns in strikes.raw to match strikes table in database
strikes.raw <- strikes.raw[,c("flight_ids", "impact", "damage", "altitude_ft", "condition_ids")]

# Change column names to match strikes table in database
colnames(strikes.raw) <- c("fid", "impact", "damage", "altitude", "conditions")

# Write fid, impact, damage, altitude and conditions Columns into the strikes table
dbWriteTable(con, name="strikes", strikes.raw, append=TRUE, row.names = FALSE)

# Store updated table after import
strikes.updated <- dbGetQuery(con, "SELECT * FROM strikes")
```

## Displaying updated tables

```{sql Show airports, echo = FALSE, eval = FALSE}
SELECT *
FROM airports
LIMIT 10;
```
```{sql Show flights, echo = FALSE, eval = FALSE}
SELECT *
FROM flights
LIMIT 10;
```
```{sql Show conditions, echo = FALSE, eval = FALSE}
SELECT *
FROM conditions
LIMIT 10;
```
```{sql Show strikes, echo = FALSE, eval = FALSE}
SELECT *
FROM strikes
LIMIT 10;
```

## Top Airports with Strikes

```{sql State vs NumStrikes, echo = FALSE}
-- Not including airportState values which are "N/A"
SELECT a.airportState, COUNT(s.sid) AS numIncidents
FROM strikes AS s
JOIN flights AS f
ON s.fid = f.fid
JOIN airports AS a
ON f.originAirport = a.aid
WHERE a.airportState <> "N/A"
GROUP BY a.airportState
ORDER BY numIncidents DESC
LIMIT 10;
```

## Analysis by Airline

```{sql Analysis by Airline, echo = FALSE}
WITH strike_count AS (
    SELECT f.airlineName, COUNT(s.sid) AS num_strikes
    FROM strikes AS s
    JOIN flights AS f
    ON s.fid = f.fid
    GROUP BY f.airlineName
)
SELECT airlineName, num_strikes AS numIncidents
FROM strike_count
WHERE num_strikes > (SELECT AVG(num_strikes) FROM strike_count)
ORDER BY num_strikes DESC;
```

## Analysis by Month

```{r Analysis by Month, echo=FALSE}
# Query to select Month name and number of strikes in each month
# Not including strikes with default lower bound date "1900-01-01"
query <- "SELECT 
    CASE SUBSTR(date, 6, 2)
      WHEN '01' THEN 'January'
      WHEN '02' THEN 'February'
      WHEN '03' THEN 'March'
      WHEN '04' THEN 'April'
      WHEN '05' THEN 'May'
      WHEN '06' THEN 'June'
      WHEN '07' THEN 'July'
      WHEN '08' THEN 'August'
      WHEN '09' THEN 'September'
      WHEN '10' THEN 'October'
      WHEN '11' THEN 'November'
      WHEN '12' THEN 'December'
    END AS month,
    COUNT(*) AS numIncidents
FROM 
strikes AS s
JOIN flights AS f
ON s.fid = f.fid
WHERE date <> '1900-01-01'
GROUP BY 
    SUBSTR(date, 6, 2)
ORDER BY 
    SUBSTR(date, 6, 2);"

# Storing result of query
result <- dbGetQuery(con, query)

# Printing first 6 rows of result
kable(head(result), format = "markdown", caption = "Displaying records 1 - 6")
```

## Trend by Month

```{r Trend by Month, echo = FALSE}
# Bar Plot for Number of Strikes by Month

# Make label text perpendicular to axis
par(las=2) 

# Store limits of y-axis (0 to 1.2 times the max value in num_strikes) into a variable
y_lim <- c(0, max(result$numIncidents) * 1.2)

# Plot month vs number of strikes with corresponding limits and color
bp <- barplot(result$numIncidents, names.arg = result$month, axis.lty = 1, ylim = y_lim, col = "blue")

# Add plot title and axis labels
title(main="Number of Incidents by Month", xlab="Month", ylab="Count")

# Add text for number of strikes at the top of each bar
text(bp, result$numIncidents, labels = result$numIncidents, pos = 1, cex = 0.8, col="white")

```

## Create Stored Procedure

```{r Create Stored Procedure, echo = FALSE, results='hide'}
# Delete audit_log table if already exists
# For testing purpose, comment if running procedure multiple times
dbExecute(con, "DROP TABLE IF EXISTS audit_log")

# SQL Query for Create audit_log table if it doesn't exist
create_audit_log_query <- "CREATE TABLE IF NOT EXISTS audit_log (
                           log_id INT AUTO_INCREMENT PRIMARY KEY,
                           modification_type VARCHAR(50) NOT NULL,
                           table_name VARCHAR(50) NOT NULL,
                           date_time DATETIME NOT NULL DEFAULT CURRENT_TIMESTAMP,
                           primary_key INTEGER NOT NULL
                        );"

# Executing query
dbExecute(con, create_audit_log_query)

# Delete stored procedure if already exists
dbExecute(con, "DROP PROCEDURE IF EXISTS remove_strike")

# SQL Query to create stored procedure
create_proc_query <- "CREATE PROCEDURE IF NOT EXISTS remove_strike(IN strike_pk INT)
                      BEGIN
                          DELETE FROM strikes WHERE sid = strike_pk;
                          INSERT INTO audit_log (modification_type, table_name, date_time, primary_key)
                          VALUES ('removal', 'strikes', CURRENT_TIMESTAMP(), strike_pk);
                      END"

# Execute SQL Query
dbExecute(con, create_proc_query)
```
## Test Stored Procedure

```{r Test Stored Procedure, eval=FALSE}
# Call stored procedure
dbExecute(con, "CALL remove_strike(2);")

# Show updated strikes table
dbGetQuery(con, "SELECT * FROM strikes LIMIT 10;")

# Show updated audit_log table
dbGetQuery(con, "SELECT * FROM audit_log;")
```

## Disconnect from Database

```{r Disconnecting from DB, results='hide', echo = FALSE, warning=FALSE, message=FALSE}
dbDisconnect(con)
```

